---
title: "Combine Irrigation Data"
author: "Mark Barrett"
format: html
---

# Libraries

```{r}
library(tidyverse)
library(here)
library(skimr)
```
# Data

Downloaded all files from the [Illinois Data Bank](https://doi.org/10.13012/B2IDB-4607538_V1).
Select all files, click *Get Custom Zip and Download Link for Selected*, unzip the file and copy all the csv files into a folder named `data`. I put the files in a subfolder named `raw` because I expect I will want to output a processed file and it's useful to differentiate earlier on. 

```{r}
data_folder = here("data", "raw")
(data_files = list.files(data_folder, full.names = T))
```

## Load

Read one file in for each data type.

According to the readme:

>All csv files in this folder specify the type of water: where GWA is Groundwater Abstractions, GWD is Groundwater Depletion, and SW is Surface water Withdrawals; followed by the year, ranging from 2008 to 2020. 

```{r}
gwa_example = str_subset(data_files, "gwa") |> 
  sample(1) |> 
  read_csv()

glimpse(gwa_example)
```

```{r}
gwd_example = str_subset(data_files, "gwd") |> 
  sample(1) |> 
  read_csv()

glimpse(gwd_example)
```

```{r}
sw_example = str_subset(data_files, "sw") |> 
  sample(1) |> 
  read_csv()

glimpse(sw_example)
```

## Clean

Pivot the data to longer format and then split the column names into the data they contain: water type, crop, and year

```{r}
gwa_example |> 
  pivot_longer(-GEOID, 
               names_to = c("measure", "crop", "year"), 
               names_sep = "\\.")
```

```{r}
gwd_example |> 
  pivot_longer(-GEOID, 
               names_to = c("measure", "crop", "year"), 
               names_sep = "\\.")
```

## Combine data

This data isn't "tidy", meaning it has information baked into column names that should* be in rows: the crop, year, and water use measure.



```{r}
gwa_files = str_subset(data_files, "gwa")
gwd_files = str_subset(data_files, "gwd")
sw_files = str_subset(data_files, "sw")

gwa_all = map(gwa_files, read_csv, col_types = "innnnnnnnnnnnnnnnnnnn", show_col_types = F) |> 
  map(pivot_longer, -GEOID, names_to = c(".value", "crop", "year"), names_sep = "\\.") |> 
  list_rbind() |> 
  # Trim leading zeroes in GEO_ID
  mutate(GEOID = str_remove_all(GEOID, "^0"))

gwd_all = map(gwd_files, read_csv, col_types = "innnnnnnnnnnnnnnnnnnn", show_col_types = F) |> 
  map(pivot_longer, -GEOID, names_to = c(".value", "crop", "year"), names_sep = "\\.") |> 
  list_rbind() |> 
  # Trim leading zeroes in GEO_ID
  mutate(GEOID = str_remove_all(GEOID, "^0"))

sw_all = map(sw_files, read_csv, col_types = "innnnnnnnnnnnnnnnnnnn", show_col_types = F) |> 
  map(pivot_longer, -GEOID, names_to = c(".value", "crop", "year"), names_sep = "\\.") |> 
  list_rbind() |> 
  # Trim leading zeroes in GEO_ID
  mutate(GEOID = str_remove_all(GEOID, "^0"))

glimpse(gwa_all)
```

Combine files

```{r}
irrigation = list(
  gwa_all, gwd_all, sw_all
) |> 
  reduce(full_join, by = c("GEOID", "crop", "year")) |> 
  select(crop, year, GEOID, sw, gwa, gwd)

glimpse(irrigation)


```

```{r}
skim(irrigation)
```

What are GWA, GWD, and SW measured in? km^3

```{r}
irrigation2 = map(data_files, read_csv, col_types = "cnnnnnnnnnnnnnnnnnnnn", show_col_types = F) |> 
  map(mutate, GEOID = str_pad(GEOID, pad = "0", width = 5, side = "left"))  |> 
    map(pivot_longer, -GEOID, 
        names_to = c("measure", "crop", "year"), 
        names_sep = "\\.", 
        names_transform = list(year = ~as.integer(.))) |> 
  list_rbind() |> 
  pivot_wider(names_from = measure, values_from = value) |> 
    select(crop, year, GEOID, sw, gwa, gwd) |> 
  mutate(gww_sus = gwa - gwd)

summary(irrigation2)
```



# Summary

Water use measures by year

```{r}
(iwu_by_year = irrigation2 |> 
  group_by(year) |> 
  summarise(across(where(is_double),sum)))
```

Avg annual IWU

```{r}
iwu_by_year |> 
  summarise(across(where(is_double), \(x) mean(x, na.rm=T)))
```

Water use measures by crop and year

```{r}
(iwu_by_cropyear = irrigation2 |> 
  group_by(year, crop) |> 
  summarise(across(where(is_double),sum)) |> 
   ungroup())
```

Avg. Annual IWU, by crop

```{r}

avg_annual_iwu = iwu_by_cropyear |> 
  group_by(crop) |> 
  summarise(across(where(is_double), \(x) mean(x, na.rm=T)))|> 
  mutate(across(where(is_double), \(x) round(x, 2))) # |>
  # mutate(across(where(is_double), \(x) scales::percent(x/sum(x))))

avg_annual_iwu
```

But the values for surface water withdrawals in the paper describing this data differ from these...

```{r}
annual_avg_crop_sww = tribble(
  ~crop, ~sw_table,
  "barley", 3.05,
  "corn", 5.79,
  "cotton",  1.76,
  "millet", .06,
  "oats", .66,
  "other_sctg2", .59,
  "other_sctg3", 9.39,
  "other_sctg4", 32.93,
  "peanuts", .06,
  "potatoes", 1.03,
  "pulses", .58,
  "rapeseed", .03,
  "rice", 2.73,
  "rye", .08,
  "sorghum", .39,
  "soybeans", .38,
  "sugarbeets", .41,
  "sunflower", .08,
  "sweetpotatoes", .01,
  "wheat", 7.35
)

avg_annual_iwu |> 
  select(crop, sw) |> 
  left_join(annual_avg_crop_sww, by = "crop") |> 
  mutate(diff = sw - sw_table)
```

Top water usage by crop in 2020

```{r}
top_crop_2020 = iwu_by_cropyear |> 
  filter(year == 2020) |> 
  arrange(desc(gwa)) |> 
  mutate(crop_rank_2020 = rank(desc(gwa))) |> 
  select(starts_with("crop"))
```


## Crop Water Usage over Time

```{r}
crop_pal = c("barley", "corn", "cotton", "Other", "Other Animal Feed", "Other Produce", "rice", "wheat")

iwu_by_cropyear |> 
  left_join(top_crop_2020, by = "crop") |> 
  mutate(crop = case_when(crop == "other_sctg2" ~ "Other Grains",
                          crop == "other_sctg3" ~ "Other Produce",
                          crop == "other_sctg4" ~ "Other Animal Feed",
                          T ~ crop),
         crop_cat = if_else(crop_rank_2020 < 8, crop, "Other")) |> 
  ggplot(aes(x = year, y = gwa))+
  geom_line(aes(group = crop, colour = crop_cat, alpha = (crop_cat != "Other")))+
  scale_colour_manual(values = )+
  scale_col
  theme_minimal()
```

