---
title: "Combine Irrigation Data"
author: "Mark Barrett"
format: 
  html:
    df-print: paged
    self-contained: true
    toc: true
    theme: 
      light: flatly
      dark: darkly
---

# Libraries

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(skimr) # For getting data summary
library(janitor) # For cleaning colnames
library(tidycensus) # For FIPS code data
library(mokeR)
```

# Data

I downloaded the files from [Illinois Data Bank](https://doi.org/10.13012/B2IDB-4607538_V1).
To follow along and copy my file organization, select all files, click *Get Custom Zip and Download Link for Selected*, unzip the file and move the csv files into a folder named `data`. I actually put the files in a subfolder named `raw` because I will want to output a processed data file and it's useful to separate raw and processed data. You can leave this step out and remove the second argument (`"raw"`) in the `here()` function below.

```{r}
data_folder = here("data", "raw")
# List the files in our data folder
(data_files = list.files(data_folder, full.names = T, pattern = "(sw|gwa|gwd)_20[0-9]{2}"))
```

## IWU Examples

Read in the 2020 file in for each data type.

```{r}
gwa_example_file = str_subset(data_files, "gwa_2020")
gwd_example_file = str_subset(data_files, "gwd_2020")
sw_example_file = str_subset(data_files, "sw_2020")

example_files = list(
  gwa_example_file, gwd_example_file, sw_example_file
  )
```


According to the readme:

>All csv files in this folder specify the type of water: where GWA is Groundwater Abstractions, GWD is Groundwater Depletion, and SW is Surface water Withdrawals; followed by the year, ranging from 2008 to 2020. 

### Cleaning

In each case, I specify the column types to ensure the data types are consistent.
I noticed FIPS codes (`GEOID`) were not being read in consistently across each file. 
Some have leading zeroes for a number of counties which results in the data being read in as 4 digits instead of 5.
To ensure the codes are consistent across all files, I pad the string up to 5 characters with 0s using `str_pad()`.

```{r}
(examples_clean = map(example_files, read_csv, 
                      show_col_types = F, col_types = "cdddddddddddddddddddd") |> 
    # Pad FIPs codes
    map(mutate, GEOID = str_pad(GEOID, pad = "0", width = 5, side = "left")))

```

### Pivot long and wide

This data isn't "tidy", meaning it has information baked into column names that should\* be in rows: crop, year, and water use measure.
To tidy this dataset such that we have crop and year in their own columns and the different measures either in a `measure` column, pivot the data to longer format and split the columns into the data contained in their name: water type, crop, and year. 

```{r}
examples_wide_list = examples_clean |> 
  map(pivot_longer, -GEOID, 
               names_to = c("measure", "crop", "year"), 
               names_sep = "\\.")

(examples_long_df = list_rbind(examples_wide_list))

```

This kind of format is useful when we want to compare measures as different colours in ggplot. 
Often it's more valuable for performing calculations to have the different measures in their own column.
To achieve this, we pivot multiple measures wider.

```{r}
(examples_wide_df = pivot_wider(examples_long_df, 
                                id_cols = c(GEOID, crop, year), 
                                names_from = measure, values_from = value))
```


## Load full IWU data

Now I'll repeat the steps used for the example files with the full data.
Load, clean, and pivot the data to long format.

```{r}
(iwu_long = map(data_files, read_csv, 
                      show_col_types = F, col_types = "cdddddddddddddddddddd") |> 
    # Pad FIPs codes
    map(mutate, GEOID = str_pad(GEOID, pad = "0", width = 5, side = "left")) |> 
   # Pivot measures
   map(pivot_longer, -GEOID, 
       names_to = c("measure", "crop", "year"), 
       names_sep = "\\.", 
       names_transform = list(year = ~as.integer(.))) |> 
   list_rbind()) |> 
  glimpse()

```

Pivot the data out to wide format.

```{r}
(iwu_wide = iwu_long |> 
  pivot_wider(names_from = measure, values_from = value) |>
  select(year, GEOID, crop, sw, gwa, gwd) |> 
    # Calculate Sustainable GWW/GWA
    mutate(sus_gwa = gwa - gwd,
           total = sw+gwa)) |> 
  glimpse()
```


```{r}
skim(iwu_wide)
```

## FIPS Codes

Join in state information based on the GEOID using the `fips_codes` dataset in `{tidycensus}`.

```{r}
library(tidycensus)
data("fips_codes")
fips_clean = fips_codes |> 
  unite(col = fips, c(state_code, county_code), sep = "")

iwu_wide_st = iwu_wide |> 
  left_join(fips_clean, by = c("GEOID" = "fips"))
```

## USGS Water Use

https://www.sciencebase.gov/catalog/item/get/5af3311be4b0da30c1b245d8

>This dataset contains water-use estimates for 2015 that are aggregated to the county level in the United States. The U.S. Geological Survey's (USGS's) National Water Use Science Project is responsible for compiling and disseminating the Nation's water-use data. Working in cooperation with local, State, and Federal agencies, the USGS has published an estimate of water use in the United States every 5 years, beginning in 1950. Water-use estimates aggregated to the State level are presented in USGS Circular 1441, "Estimated Use of Water in the United States in 2015" (Dieter and others, 2018). This dataset contains the county-level water-use data that support the state-level estimates in Dieter and others 2018. This dataset contains data for public supply, domestic, irrigation, thermoelectric power, industrial, mining, livestock, and aquaculture water-use categories.

```{r}
usgs_file = list.files(data_folder, pattern = "usco.+\\.csv$", full.names = T)

(usgs = read_csv(usgs_file, skip = 1, na = "--") |> 
    clean_names()) |> 
  glimpse()

usgs_iwu_long = usgs |> 
  # Select groundwater, surface, and total freshwater use measures
  select(state:year, contains("_w_fr_to"), contains("_wgw_fr"),
         contains("_wsw_fr"), contains("_rec_ww")) |> 
  # Rename the measures
  rename_with(\(x) str_replace_all(x, c("(?<=_)w_fr_to" = "total",
                                        "(?<=_)wgw_fr" = "gwa",
                                        "(?<=_)wsw_fr" = "sw",
                                        "(?<=_)rec_ww" = "reclaimed",
                                        "(?<=_)c_us_fr" = "consumption")),
              -c(state:year))

(usgs_iwu_wide = usgs_iwu_long |> 
  # Pivot the use categories to a column and separate out the different measures
  pivot_longer(cols = contains("_"), names_to = c("category", ".value"), names_sep = "_",
               names_transform = list(category = \(x) factor(x, 
                                                             levels = c("ps", "do", "in", "ir", 
                                                                        "ic", "ig", "li", "aq", 
                                                                        "mi", "pt", "po", 
                                                                        "pc", "to"),
                                                             labels = c("Public Supply", "Domestic", "Industrial", "Irrigation", 
                                                                        "Crop Irrigation", "Golf Irrigation", "Livestock", "Aquaculture", 
                                                                        "Mining", "Thermoelectric", "Thermoelectric (Once-through)", 
                                                                        "Thermoelectric (Recirculating)", "Total"))))) |> 
  glimpse()

```
These irrigation water use measures are in millions of gallons per day. 
Let's see if we can convert this to km^3 per year.

```{r}

mgal_to_km3 = function(x){
  # Multiple by days to get gallons/year
  mgal_year = x*365.25
  # Multiply by 0.0000037854118 because this is the conversion factor for mgal to km^3?
  mgal_year*0.000003785411784
}
usgs_iwu_wide |> 
  # filter(category == "Crop Irrigation") |> 
  group_by(category) |> 
  # mutate(across(-c(state:year), mgal_to_km3)) |>
  summarise(across(-c(state:year), \(x) sum(x, na.rm=T)))


```
As we'll see next, the Crop Irrigation estimates match that of our IWU estimates for 2015.


# Summary

## Annual Irrigation levels, by Source

Based on these data, how much water are used for irrigating crops each year?

```{r}
(iwu_total_annual = iwu_wide_st |> 
  group_by(year) |> 
  summarise(across(where(is_double),sum)))
```

```{r}
annual_iwu = iwu_long |> 
  group_by(year, measure) |> 
  summarise(value = sum(value), .groups = "drop") |> 
  mutate(measure = factor(measure, levels = c("sw", "gwa", "gwd"), 
                          labels = c("Surface", "Groundwater", "Groundwater Depletion")))

# filter(annual_iwu_sus, measure != "Groundwater Withdrawals") |> 
filter(annual_iwu, measure != "Groundwater Depletion") |> 
  ggplot(aes(x = year, y = value))+
  geom_area(aes(fill = measure), colour = "grey29", linewidth = .2, alpha = .98, show.legend = F)+
  # geom_line(data = filter(annual_iwu, measure == "Groundwater Withdrawals"), colour = "black")+
  scale_x_continuous(name = NULL, breaks = scales::pretty_breaks(), expand = expansion(mult = c(0, .025)))+
  scale_y_continuous(name = "Irrigation Water Use (km^3)", breaks = scales::pretty_breaks(n = 6), expand = expansion(mult = c(0, .15)))+
  scale_fill_manual(name = NULL, values = c("mediumseagreen", "cornflowerblue", "orangered"))+
  labs(title = "How much water is used to irrigate crops?", 
       subtitle = "Surface and Groundwater Withdrawals for Irrigation Use, 2008-2020",
       caption = source_caption("Ruess, P.J., Konar, M., Wanders, N. , & Bierkens, M. (2023)"))+
  annotate("text", x = 2014, y = 75, label = "Surface", colour = "#FFFFFF", size = 5, )+
  annotate("text", x = 2014, y = 25, label = "Groundwater", colour = "#FFFFFF", size = 5)+
  theme_moke(plots_pane = T)
```




What is the yearly average for each IWU measure between 2008-2020?

```{r}
iwu_total_annual |> 
  summarise(across(where(is_double), \(x) mean(x, na.rm=T)))
```


Calculate yearly crop IWUs

```{r}
(iwu_by_cropyear = iwu_wide |> 
  group_by(year, crop) |> 
  summarise(across(where(is_double),sum)) |> 
   ungroup())
```

What is each crop's yearly average for the IWU measures between 2008-2020?

```{r}
(avg_annual_iwu = iwu_by_cropyear |> 
  group_by(crop) |> 
  summarise(across(where(is_double), \(x) mean(x, na.rm=T)))|> 
  mutate(across(where(is_double), \(x) round(x, 2))))
```

But the values I calculatefor surface water withdrawals differ from [Table 2 in the authors' paper describing this data](https://agupubs.onlinelibrary.wiley.com/doi/10.1029/2022WR032804#wrcr26385-tbl-0002)

```{r}
annual_avg_crop_sww = tribble(
  ~crop, ~sw_table,
  "barley", 3.05,
  "corn", 5.79,
  "cotton",  1.76,
  "millet", .06,
  "oats", .66,
  "other_sctg2", .59,
  "other_sctg3", 9.39,
  "other_sctg4", 32.93,
  "peanuts", .06,
  "potatoes", 1.03,
  "pulses", .58,
  "rapeseed", .03,
  "rice", 2.73,
  "rye", .08,
  "sorghum", .39,
  "soybeans", .38,
  "sugarbeets", .41,
  "sunflower", .08,
  "sweetpotatoes", .01,
  "wheat", 7.35
)

avg_annual_iwu |> 
  select(crop, sw) |> 
  left_join(annual_avg_crop_sww, by = "crop") |> 
  mutate(diff = sw - sw_table)
```

Top water usage by crop in 2020

```{r}
top_crop_2020 = iwu_by_cropyear |> 
  filter(year == 2020) |> 
  arrange(desc(gwa)) |> 
  mutate(crop_rank_2020 = rank(desc(gwa))) |> 
  select(starts_with("crop"))
```

# Visualization Roughwork

What % of total water use does groundwater supply? How does this compare to 1950?

```{r}
iwu_total_annual |> 
  mutate(gwa_pct = gwa/total) |> 
  ggplot(aes(x = year, y = gwa_pct))+
  geom_line()+
  geom_hline(yintercept =.23, linetype = "dashed")+
  scale_y_continuous(limits = c(0, NA), labels = scales::percent)+
  theme_minimal()
```

Surface vs groundwater withdrawals as an area chart

```{r}
annual_iwu_sus = iwu_wide |> 
  pivot_longer(cols = c(sw, gwa, gwd, sus_gwa), names_to = "measure", values_to = "value") |> 
  group_by(year, measure) |> 
  summarise(value = sum(value), .groups = "drop") |> 
  mutate(measure = factor(measure, levels = c("sw", "gwa", "sus_gwa", "gwd"), 
                          labels = c("Surface Water Withdrawals", "Groundwater Withdrawals", str_wrap("Groundwater Withdrawals (Sustainable)", 32), "Groundwater Depletion")))


```
How much of the groundwater withdrawals are sustainable vs depletion?

```{r}
filter(annual_iwu_sus, measure != "Groundwater Withdrawals") |> 
  # mutate(value = if_else(measure == "Surface Water Withdrawals", value, -value)) |> 
  ggplot(aes(x = year, y = value))+
  geom_area(aes(fill = measure), colour = "grey29", linewidth = .2, alpha = .9)+
  # geom_area(data = filter(annual_iwu, measure == "Groundwater Depletion"), alpha = .7, aes(fill = measure), linetype = "dashed")+
  scale_x_continuous(name = NULL, breaks = scales::pretty_breaks(), expand = expansion(add = .5))+
  scale_y_continuous(name = "Irrigation Water Use (km^3)", expand = expansion())+
  # scale_colour_manual(name = NULL, values = "orangered")+
  scale_fill_manual(name = NULL, values = c("mediumseagreen", "cornflowerblue", "orangered"))+
  labs(title = "Agricultural Irrigation Water Use", subtitle = "Sources of Irrigation Water")+
  theme_minimal()
```

## Crop Water Usage over Time

```{r}
crop_pal = c("barley", "corn", "cotton", "Other", "Other Animal Feed", "Other Produce", "rice", "wheat")

iwu_by_cropyear |> 
  left_join(top_crop_2020, by = "crop") |> 
  mutate(crop = case_when(crop == "other_sctg2" ~ "Other Grains",
                          crop == "other_sctg3" ~ "Other Produce",
                          crop == "other_sctg4" ~ "Other Animal Feed",
                          T ~ crop),
         crop_cat = if_else(crop_rank_2020 < 8, crop, "Other")) |> 
  ggplot(aes(x = year, y = gwa))+
  geom_line(aes(group = crop, colour = crop_cat, alpha = (crop_cat != "Other")))+
  # scale_colour_manual(values = )+
  # scale_col
  theme_minimal()
```

## Irrigation by State

```{r}
iwu_wide_st |> 
  group_by(state_name) |> 
  summarise(across(where(is_double), sum)) |> 
  arrange(state_name)
```

